---
title: "Kate propionate paper mouse data picrust2"
output: html_document
date: "2025-06-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# --- Working directory guard ---------------------------------------------------
# Every path below is relative to this file's own folder (scripts/), which is where
# knitr runs an Rmd from by default. If someone sources the chunks from the project
# root instead, every read silently fails partway through a long run. Check up front
# and fail with a message that says what to do.
if (!dir.exists("../data") || !file.exists("../renv.lock")) {
  stop("This notebook must be run from the 'scripts/' folder of the propionate repo.\n",
       "  Current working directory: ", normalizePath("."), "\n",
       "  In RStudio, open propionate.Rproj and knit this file; from the console, ",
       "setwd() to scripts/ first.")
}

# --- Output folders ------------------------------------------------------------
# results/ holds every figure this notebook writes. It is not tracked in git (the
# figures are regenerated from this script), so create it on a fresh clone.
dir.create("../results", showWarnings = FALSE)
# data/derived/ holds files this notebook *generates* as input for PICRUSt2. Keeping
# them out of data/ makes it obvious which files are raw inputs and which are built.
dir.create("../data/derived", showWarnings = FALSE)

# --- Packages ------------------------------------------------------------------
# Loaded up front so a missing dependency fails in the first second rather than
# twenty minutes into the run. Exact versions are pinned in renv.lock.
library(tidyverse)
library(readxl)      # reads the GVHD clinical scoring spreadsheet
library(ggrepel)
library(Biostrings)  # FASTA handling for the PICRUSt2 input prep
library(lmerTest)    # mixed-effects models that report p-values
library(DESeq2)      # differential abundance of ASVs
library(vegan)       # PERMANOVA (adonis2)

# --- Reproducibility -----------------------------------------------------------
# adonis2() below is a permutation test: without a fixed seed its p-value changes on
# every run, so the numbers in the manuscript would not be reproducible. Seed once
# here for a top-to-bottom knit, and again immediately before the test itself so
# that re-running that chunk alone gives the same answer.
set.seed(20250602)
```

**Run order matters.** This notebook is written to be knitted, or run top to bottom in
a fresh session. Several object names (`data_full`, `plot_data`, `meta`, `tax_raw`,
`top_pathways`) are deliberately reused between the steady-state section and the
transplant section, so running chunks out of order can silently analyze the wrong
dataset. The chunks that feed manuscript figures assert which dataset they are
holding before they compute anything.

renv reminders, for whoever picks this up next:

- `renv::restore()` recreates the exact package library recorded in `renv.lock`. Run
  this first on a new machine — see the README for the required R version.
- `renv::snapshot()` writes any newly installed packages back into `renv.lock`. Run
  it whenever you add a `library()` call, or the next person's restore will not have
  what the script needs.
- `renv::status()` reports drift between the lockfile and what is actually installed.


# Preparing the PICRUSt2 inputs

These two chunks convert the Zymo ASV output into the formats PICRUSt2 expects. They
are set to `eval=FALSE` on purpose: PICRUSt2 itself was run separately on Linux, and
its outputs are already shipped in `data/picrust2_out_*`, so nothing downstream in
this notebook depends on re-running them. They are kept for provenance — to show
exactly how the PICRUSt2 inputs were built — and both now write into `data/derived/`
rather than overwriting files in `data/`.

PICRUSt2 was installed in a fresh conda environment with
`mamba install bioconda::picrust2`, using the older `picrust2_pipeline.py` 2.0.3-b.

## ASV count table: wide CSV to the tab-delimited layout PICRUSt2 wants

The target format is one row per ASV and one column per sample:

    #OTUID  SampleA  SampleB  SampleC
    ASV_1   150      0        210
    ASV_2   0        55       84
    ASV_3   78       12       0

```{r eval=FALSE}
# Find the raw per-experiment abundance tables shipped with the repo
asv_cts <- list.files('../data/', pattern = 'ASV_Abundance_Table.csv', full.names = T)

# Transpose each one into the PICRUSt2 layout and write it to data/derived/.
# Writing to a separate folder keeps the raw inputs in data/ untouched, so this
# chunk can be re-run any number of times without changing its own input.
asv_cts %>%
  map(function(fn){
    out_fn <- file.path('../data/derived', str_replace(basename(fn), '\\.csv$', '.tsv'))
    read_csv(fn) %>%
      # Long-then-wide pivot swaps the sample and ASV axes
      gather('seq_id','count', names(.)[2]:names(.)[ncol(.)]) %>%
      spread('seqs','count', fill = 0) %>%
      janitor::clean_names() %>%
      rename(`#ASVID` = seq_id)  %>%
      write_tsv(file = out_fn)
    out_fn
  })
```

## ASV sequences: strip the size annotation from the FASTA headers

PICRUSt2 matches sequence names against the count table, so the trailing
` size=NNN` that the Zymo pipeline appends has to go.

```{r eval=FALSE}
# Sanity check that the ASV ids in the count table and the FASTA agree
test <- read_csv(asv_cts[1]) %>%
      gather('seq_id','count', names(.)[2]:names(.)[ncol(.)])

fasta_file <- '../data/transplant_sv.seqs.fna'
dna_sequences <- readDNAStringSet(fasta_file)

# Drop the " size=..." suffix that Zymo adds to each header
cleaned_names <- sub(" size=.*", "", names(dna_sequences))
names(dna_sequences) <- cleaned_names

# Confirm the cleaned names actually match the count table before writing anything
stopifnot(length(intersect(cleaned_names, test$seq_id)) > 0)

# Write the cleaned FASTA to data/derived/ rather than back over the input file.
# The original version of this chunk wrote to `fasta_file` itself, which meant every
# run mutated a tracked input and the second run operated on different data.
writeXStringSet(dna_sequences,
                filepath = '../data/derived/transplant_sv.seqs.cleaned.fna',
                format = "fasta")
```

# pathway output 

```{r}
pw_fns <- c('../data/picrust2_out_steady/pathways_out/path_abun_unstrat.tsv.gz','../data/picrust2_out_transplant//pathways_out/path_abun_unstrat.tsv.gz')

# take a look at the total 

pw_dfs <- pw_fns %>% 
  set_names(c('steady','transplant')) %>% 
  map(function(fn){
    df = read_tsv(fn)
    return(df)
  })
```


```{r}
combined_pathway_data <- pw_dfs %>%
  purrr::map_dfr(~ .x %>%
                   pivot_longer(
                     cols = -pathway,        # Gathers all columns except 'pathway'
                     names_to = "sample_id", # New column for original column names (sample IDs)
                     values_to = "abundance" # New column for the values (abundances)
                   ),
                 .id = "exp")
 
#  Calculate the total abundance for each unique pathway across all samples
total_pathway_abundance <- combined_pathway_data %>%
  group_by(pathway) %>%
  summarise(total_abundance = sum(abundance, na.rm = TRUE)) %>% # Sum abundances for each pathway
  ungroup() # Ungroup for subsequent operations if any

#  Visualize the distribution of total pathway abundances using a histogram

ggplot(total_pathway_abundance, aes(x = total_abundance)) +
  geom_histogram(binwidth = 100000, # Adjust binwidth based on the data's range
                 fill = "#2196F3",   # A nice blue color
                 color = "white",    # White border for bars
                 alpha = 0.8,        # Slightly transparent
                 boundary = 0) +     # Align bins to start at 0
  labs(
    title = "Distribution of Total Predicted Pathway Abundances",
    x = "Total Abundance",
    y = "Number of Pathways"
  ) +
  theme_minimal() + # Use a minimalist theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), # Center and style title
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid.minor = element_blank(), # Remove minor grid lines
    panel.background = element_rect(fill = "white", colour = "white"), # White background
    plot.background = element_rect(fill = "white", colour = "white") # White plot background
  )
```
```{r}
# to look at the total abundance of all pathways in each sample 
total_sample_abundance <- combined_pathway_data %>%
  group_by(sample_id, exp) %>% # Group by sample_id and source_df 
  summarise(total_abundance = sum(abundance, na.rm = TRUE)) %>% # Sum abundances for each sample
  ungroup() # Ungroup for subsequent operations if any

#  Visualize the distribution of total sample abundances using a histogram

ggplot(total_sample_abundance, aes(x = total_abundance)) +
  geom_histogram(binwidth = 1000000, 
                 fill = "#00B0FF",   # A brighter blue color
                 color = "white",    # White border for bars
                 alpha = 0.8,        # Slightly transparent
                 boundary = 0) +     # Align bins to start at 0
  labs(
    title = "Distribution of Total Pathway Abundances per Sample",
    x = "Total Abundance (Sum of all Pathways)",
    y = "Number of Samples"
  ) +
  theme_minimal() + # Use a minimalist theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), # Center and style title
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid.minor = element_blank(), # Remove minor grid lines
    panel.background = element_rect(fill = "white", colour = "white"), # White background
    plot.background = element_rect(fill = "white", colour = "white") # White plot background
  )


```

```{r}
#  Visualize the total abundance of all pathways in each sample as bars,
ggplot(total_sample_abundance, aes(x = forcats::fct_reorder(sample_id, sample_id, .fun = function(x) as.numeric(factor(x))), # Sorts alphabetically
                                   y = total_abundance,
                                   fill = exp)) +
  geom_col(color = "white", alpha = 0.8) + # Use geom_col for bar plots, with white borders
  labs(
    title = "Total Pathway Abundance per Sample",
    x = "Sample ID (Sorted Alphabetically)",
    y = "Total Abundance (Sum of all Pathways)",
    fill = "Exp" # Legend title for fill color
  ) +
  theme_minimal() + # Use a minimalist theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), # Center and style title
    axis.title = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8), # Rotate x-axis labels for readability
    axis.text.y = element_text(size = 10),
    legend.position = "bottom", # Place legend at the bottom
    panel.grid.minor = element_blank(), # Remove minor grid lines
    panel.background = element_rect(fill = "white", colour = "white"), # White background
    plot.background = element_rect(fill = "white", colour = "white") # White plot background
  ) +
  scale_fill_brewer(palette = "Set1") + # Use a colorblind-friendly palette # Use a colorblind-friendly palette
  facet_wrap(~ exp, nrow = 1, scales = "free_x")


```

```{r}
# look at the sampleids which is relavent to the study design
samples <- total_sample_abundance %>% distinct(exp, sample_id)
```

## steady state experiment

MaAsLin2 is built for this kind of analysis. It can use the full dataset (pre and post) and properly model the paired samples by treating mouse_id as a random effect, giving a more accurate and powerful results.

```{r}
sample_names <- samples %>% filter(exp == 'steady') %>% pull(sample_id)

metadata <- tibble(SampleID = sample_names) %>%
  # Separate the sample ID string into four parts based on the underscore
  separate(
    SampleID,
    into = c("Treatment", "Timepoint", "MouseWord", "MouseNum"),
    sep = "_",        # <-- The separator is now an underscore
    remove = FALSE    # Keep the original SampleID column
  ) %>%
  # Combine 'MouseWord' and 'MouseNum' to create a unique ID for each mouse
  mutate(MouseID = paste(MouseWord, MouseNum, sep = "_")) %>%
  # Select the final, clean columns we need for the analysis
  select(SampleID, Treatment, Timepoint, MouseID) %>% 
  mutate(Treatment = factor(Treatment, levels = c('pbs','propionate')),
         Timepoint = factor(Timepoint, levels = c('pre','post')) )
```


    
### the LME looking at the unstratified data

- Why the Mixed-Effects Model is Still the Best Choice

- Fixed Effects:  the variables I want to test: time (pre vs. post) and treatment_group (PBS vs. Propionate). The model will test if the change over time is different between the two groups (the time * treatment_group interaction).

- Random Effect: This is what makes the model perfect for the data. By adding mouse_id as a random effect, you are telling the model that the two measurements from Mouse_1 are related to each other, and not independent like the measurements from Mouse_2. It correctly pairs the pre- and post-treatment samples for each mouse. 

```{r}
# Load pathway abundance data (from path_abun_unstrat.tsv)
path_abun <- read.table("../data/picrust2_out_steady/pathways_out/path_abun_unstrat.tsv.gz", 
                        header = TRUE, 
                        sep = "\t", 
                        row.names = 1, 
                        check.names = FALSE)

path_abun_relative <- as.data.frame(prop.table(as.matrix(path_abun), margin = 2))

# Scale to Counts Per Million (CPM) for better log-transformation behavior
path_abun_cpm <- path_abun_relative * 1000000

# Prepare and Merge the Data

path_abun_t <- as.data.frame(t(path_abun_cpm))

# Add a SampleID column from the row names
path_abun_t$SampleID <- rownames(path_abun_t)

# Pivot the abundance data from wide to long format
path_abun_long <- path_abun_t %>%
  pivot_longer(
    cols = -SampleID, # Pivot all columns except SampleID
    names_to = "Pathway",
    values_to = "CPM"
  )

# Now, merge the long abundance data with the metadata
# This creates one master data frame for analysis
data_full <- left_join(path_abun_long, metadata, by = "SampleID") 

#  TRANSFORM and RUN LME model

data_full$log_cpm <- log2(data_full$CPM + 1)
```


```{r}
# Group by pathway and run the LME model on the transformed data
lme_results <- data_full %>%
  group_by(Pathway) %>%
  do({
    model <- lmer(log_cpm ~ Timepoint * Treatment + (1 | MouseID), data = .)
    coeffs <- as.data.frame(summary(model)$coefficients)
    interaction_row <- coeffs[rownames(coeffs) == "Timepointpost:Treatmentpropionate", ]
    data.frame(
      Estimate = interaction_row$Estimate,# The Estimate is a log fold change because we analyzed log-transformed data
      P_value = interaction_row$`Pr(>|t|)`
    )
  }) %>%
  ungroup()
# boundary (singular) fit: see help('isSingular'): This warning means the model found that for certain pathways, the variation between individual mice was zero. This is common for rare or low-variance pathways. It is not a critical error, and the results for your main question (the effect of Timepoint * Treatment) are still considered reliable.

# Adjust P-values
lme_results_final <- lme_results %>%
  mutate(FDR = p.adjust(P_value, method = "BH")) %>%
  arrange(FDR)

# View the final, correct results
head(lme_results_final)

```

```{r}
# Define thresholds for significance and effect size
fdr_threshold <- 0.05
estimate_threshold <- 2.0 # You can adjust this threshold for labeling

plot_data <- lme_results_final %>%
  mutate(
    # Create a column for coloring points based on significance and direction
    significance = case_when(
      FDR < fdr_threshold & Estimate > estimate_threshold  ~ "Increased",
      FDR < fdr_threshold & Estimate < -estimate_threshold ~ "Decreased",
      TRUE                                             ~ "Not Significant"
    ),
    # Create a column for labeling only the most interesting points
    label = ifelse(FDR < fdr_threshold & abs(Estimate) > estimate_threshold, Pathway, "")
  )

# Create the Volcano Plot with ggplot2
volcano_plot <- ggplot(plot_data, aes(x = Estimate, y = -log10(FDR))) +
  # Add points, colored by our new 'significance' column
  geom_point(aes(color = significance), alpha = 0.7, size = 2.5) +
  
  # Add non-overlapping text labels for the top pathways using ggrepel
  geom_text_repel(aes(label = label), 
                  box.padding = 0.5, 
                  point.padding = 0.5,
                  max.overlaps = Inf, # Allow all labels to be plotted
                  segment.color = 'grey50',
                  size = 3.5) +
  
  # Define the custom colors for our groups
  scale_color_manual(values = c("Increased" = "#d62728", 
                                "Decreased" = "#1f77b4", 
                                "Not Significant" = "grey80")) +
  
  # Add vertical lines for the effect size (Estimate) threshold
  geom_vline(xintercept = c(-estimate_threshold, estimate_threshold), linetype = "dashed", color = "grey50") +
  
  # Add a horizontal line for the FDR significance threshold
  geom_hline(yintercept = -log10(fdr_threshold), linetype = "dashed", color = "grey50") +
  
  # Add labels and titles
  labs(
    title = "Propionate Treatment vs. PBS Control",
    subtitle = "Changes in Microbial Metabolic Pathways",
    x = "Log2 Fold Change (Estimate)",
    y = "-log10(FDR)",
    color = "Significance"
  ) +
  
  # Apply a clean, publication-ready theme
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
volcano_plot
ggsave('../results/F1I.pdf', width = 8)
```


```{r}
### visualize the log2 counts of the sig pathways
top_pathways <- plot_data %>%
  filter(significance != 'Not Significant') %>% 
  arrange(desc(significance)) %>%       # Sort by FDR to get the most significant
  pull(Pathway)          # Extract the pathway names into a vector

#  Filter the main data frame for only these top pathways
plot_data <- data_full %>%
  filter(Pathway %in% top_pathways) %>%
  # Important for ordering the grid: convert Pathway to a factor based on significance
  mutate(Pathway = factor(Pathway, levels = top_pathways))

#  Create the faceted plot
# The core is the same, but we add `facet_wrap()` at the end.
faceted_interaction_plot <- ggplot(plot_data, 
                                   aes(x = Timepoint, y = log_cpm, group = MouseID, color = Treatment)) +
  
  # Add lines for each mouse
  geom_line(alpha = 0.4, size = 0.8) +
  
  # Add points for each measurement
  geom_point(alpha = 0.7, size = 1.5) +
  
  # Add a thicker line showing the AVERAGE trend for each treatment group
  stat_summary(aes(group = Treatment), fun = mean, geom = "line", size = 1.5, linetype = 'dashed') +
  
  # Use consistent colors
  scale_color_manual(values = c("pbs" = "#1f77b4", "propionate" = "#d62728")) +
  
  # *** THE KEY STEP: Create a grid of plots ***
  # `~ Pathway` tells ggplot to make a panel for each pathway.
  # `ncol = 4` arranges them into a grid with 4 columns.
  # `scales = "fixed"` ensures the Y-axis is the same for all panels, which is crucial for comparison.
  facet_wrap(~ Pathway, ncol = 6, scales = "fixed") +
  
  # Add informative labels and a title
  labs(
    title = "Significant Pathways: Pre vs. Post Treatment",
    subtitle = "Individual mouse trajectories shown, with group averages in bold",
    x = "Timepoint",
    y = "Log2(CPM + 1) Abundance",
    color = "Treatment Group"
  ) +
  
  # Apply a theme suitable for a faceted plot
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 8) # Style the facet titles
  )

#  Display the plot
ggsave('../results/01_steady_sig_samples_viz.jpg', width = 12)

```


### the stratified output

```{r}
# Load the taxonomy file exported from QIIME2
# We need to do some cleaning to separate the taxonomy string into levels
tax_raw <- read_tsv("../data/ASV_tax_assignments_steady.txt", col_names = c('taxon','Taxon','X3','X4')) 
  
tax <- tax_raw %>%
  # Separate the long taxon string into taxonomic ranks
  separate(Taxon, into = c("k", "p", "c", "o", "f", "g", "s"), sep = ";") %>%
  # Clean up the names (e.g., remove "g__")
  mutate(g = str_remove(g, "g__")) %>%
  select(taxon, Genus = g) # Keep only the Genus level for this plot

# Load the large taxon contribution file
contrib <- read_tsv('../data/picrust2_out_steady/pathways_out/path_abun_contrib.tsv.gz') %>% rename_with(~ "function_", .cols = 2)

#  Define Key Pathways and Prepare the Data ---

# Define the pathways we want to visualize
pathways_of_interest <- c(
  "PWY-7282", # Vitamin B1 Synthesis (Increased)
  "PWY-4984", # Urea Cycle / N-Recycling (Increased)
  "PWY-6263"  # Vitamin K2 Synthesis (Decreased)
)

# Filter the contribution file for our pathways and merge with taxonomy and metadata
plot_data_prep <- contrib %>%
  filter(function_ %in% pathways_of_interest) %>%
  left_join(tax, by = "taxon") %>%
  # If a taxon has no genus assigned, label it "Unassigned"
  mutate(Genus = ifelse(Genus == 'NA' | Genus == "", "Unassigned", Genus)) %>%
  left_join(metadata, by = c("sample" = "SampleID"))

#  Summarize Data for Plotting ---
# The key value, taxon_function_abun, is calculated by multiplying the raw count of a microbe by its predicted gene copy number for that function. This gives you a single number representing that microbe's total functional contribution to a pathway in your sample.

# Find the top 10 most influential genera across all samples/pathways to keep the plot clean
top_genera <- plot_data_prep %>%
  group_by(Genus) %>%
  summarise(total_contrib = sum(taxon_function_abun)) %>%
  arrange(desc(total_contrib)) %>%
  slice_head(n =10) %>%
  pull(Genus)

# Summarize the data, grouping minor genera into an "Other" category
summary_data <- plot_data_prep %>%
  # Group everything not in our top list into "Other"
  mutate(Genus = ifelse(Genus %in% top_genera, Genus, "Other")) %>%
  # Group by our experimental variables and the new Genus column
  group_by(Treatment, Timepoint, Pathway = function_, Genus) %>%
  # Calculate the total contribution for each genus in each group
  summarise(avg_contribution = mean(taxon_function_abun), .groups = 'drop') %>%
  # Set the order of Timepoint for the plot
  mutate(Timepoint = factor(Timepoint, levels = c("pre", "post")))


#  Create the Faceted Stacked Bar Plot ---

contribution_plot <- ggplot(summary_data, aes(x = Treatment, y = avg_contribution, fill = Genus)) +
  # Create the stacked bars
  geom_bar(stat = "identity", position = "stack") +
  
  # Split the plot into panels for each Timepoint (Pre vs. Post)
  facet_grid(Pathway ~ Timepoint, scales = "free_y") +
  
  # Use a color-blind friendly palette with enough colors for top genera + "Other"
  scale_fill_brewer(palette = "Paired") +
  
  # Add labels and titles
  labs(
    title = "Key Microbial Contributors to Metabolic Pathways",
    subtitle = "Comparing Pre- vs. Post-Treatment in PBS and Propionate Groups",
    x = "Treatment Group",
    y = "Average Pathway Abundance (Contribution)",
    fill = "Contributing Genus"
  ) +
   
  # Apply a clean theme
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom",
    strip.text.x = element_text(face = "bold"),
    strip.text.y = element_text(face = "bold", size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Display the plot
print(contribution_plot)

ggsave('../results/01_steady_3_pw_genus_F1J.pdf', width = 10, height = 9)
```
### "Propionate-Signature" Bacteria 

```{r}
# --- Load and Format Data ---

count_table <- read_csv("../data/steady_ASV_Abundance_Table.csv" ) %>%  column_to_rownames('seqs')
# Convert the biom object to a standard numeric matrix
count_matrix <- as.matrix(count_table) %>% t()

colnames(count_matrix) <- tolower(str_replace_all(colnames(count_matrix), "\\.", "_"))

# Load the taxonomy data
tax_raw <- read_tsv("../data/ASV_tax_assignments_steady.txt", col_names = F)
tax_clean <- tax_raw %>%
  dplyr::rename(ASV = X1) %>%
  # Separate the long taxon string into ranks for easier labeling later
  separate(X2, into = c("k", "p", "c", "o", "f", "g", "s"), sep = ";") %>%
  # Clean up the names (e.g., remove "g__")
  mutate(g = str_remove(g, "g__"),
         s = str_remove(s, "s__"),
         f = str_remove(f, "f__")) %>%
  select(ASV, Family = f, Genus = g, Species = s)

# --- Prepare Data for DESeq2 ---

# The goal is to find the signature bacteria AFTER treatment.
# So, we will filter our data to only include the 'post' samples.
meta_post <- metadata %>% 
  mutate(Timepoint = as.character(Timepoint)) %>% 
  filter(Timepoint == "post")

# The columns in the count matrix must be in the same order as the rows in the metadata.
# This reorders the count matrix columns to match the metadata.
count_matrix_post <- count_matrix[, meta_post$SampleID]

# --- Run DESeq2 Analysis ---

# Create the DESeqDataSet object. This object bundles the counts,
# metadata, and the experimental design for the analysis.
# The design `~ Treatment` tells DESeq2 to find differences based on the Treatment column.
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_post,
  colData = meta_post,
  design = ~ Treatment
)

# Run the main DESeq2 function. This performs normalization, dispersion estimation,
# and fits the model to find differentially abundant features.
dds_result <- DESeq(dds)

# Get the results for our comparison of interest: propionate vs. the PBS control.
# A positive log2FoldChange means the ASV is more abundant in the propionate group.
res <- results(dds_result, contrast = c("Treatment", "propionate", "pbs"))

# Convert the results to a tidy data frame for plotting
res_df <- as.data.frame(res) %>%
  rownames_to_column(var = "ASV") %>%
  # Add the taxonomy information to our results
  left_join(tax_clean, by = "ASV") %>%
  # Filter for statistically significant results (FDR < 0.05)
  filter(padj < 0.05) %>%
  # Sort by the fold change to see the biggest differences first
  arrange(desc(log2FoldChange))

# --- Visualize the Results ---

# Create a clean label for the plot, using Genus and ASV ID.
res_df <- res_df %>%
  mutate(PlotLabel = str_glue('f_{Family}|g_{Genus}|s_{Species} ({ASV})'))

# Create a "dot plot" of the top 20 signature ASVs
signature_plot <- ggplot(head(res_df, 20), aes(x = log2FoldChange, y = reorder(PlotLabel, log2FoldChange))) +
  geom_point(aes(color = log2FoldChange, size = -log10(padj)), alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_gradient2(low = "#1f77b4", mid = "grey", high = "#d62728", midpoint = 0) +
  labs(
    title = "Steady state: Top Propionate-Signature Bacteria (vs. PBS)",
    subtitle = "Analysis of post-treatment samples",
    x = "Log2 Fold Change (More abundant in Propionate ->)",
    y = "ASV (annotated with Genus)",
    color = "Log2 Fold Change",
    size = "Significance (-log10 FDR)"
  ) +
  theme_minimal()

# Display the plot
print(signature_plot)

ggsave('../results/F1G.pdf', width = 8)
```

## transplant experiment

```{r}
sample_names <- combined_pathway_data %>% filter(exp == 'transplant') %>% distinct(sample_id) %>% pull(sample_id)

metadata_df <- tibble(SampleID = sample_names) %>%
  # The separate() function is perfect for splitting a column by a delimiter.
  # Here, we split the 'SampleID' column into three new columns using "_" as the separator.
  separate(
    SampleID, 
    into = c("Timepoint", "Treatment", "MouseNumber"), 
    sep = "_",
    remove = FALSE # This keeps the original SampleID column, which is good practice
  ) %>%
  # Create a unique MouseID by combining the treatment group and mouse number.
  # This is important for pairing the pre/post samples correctly in a mixed-effects model.
  mutate(
    MouseID = paste(Treatment, MouseNumber, sep = "_")
  ) %>%
  # Optional: Reorder the columns for better readability
  select(SampleID, MouseID, Treatment, Timepoint, MouseNumber) %>% 
  mutate(Timepoint = factor(Timepoint, levels = c('pre','post')) )
```

```{r}
#  Load and Prepare Data ---

# Load the metadata 
meta <- metadata_df

# Load pathway abundance data (from path_abun_unstrat.tsv.gz)
path_abun <- combined_pathway_data %>% filter(exp == 'transplant')
```


```{r}
#  Normalize and Transform ---

# Convert raw counts to relative abundance, then to CPM, then log2 transform
path_abun_processed <- path_abun %>%
  group_by(sample_id) %>%
  # Normalize to relative abundance (Total Sum Scaling)
  mutate(rel_abun = abundance / sum(abundance)) %>%
  # Scale to Counts Per Million and log2 transform
  mutate(log2_cpm = log2((rel_abun * 1000000) + 1)) %>%
  select(SampleID = sample_id, Pathway = 2, log2_cpm) %>%
  ungroup()

#  Merge and Run LME Model ---

# Join the processed abundance data with the metadata
data_full <- inner_join(path_abun_processed, meta, by = "SampleID")
```


```{r}
# VERY IMPORTANT: For the primary statistical test, we only compare Propionate vs. PBS.
# The TCD group is a benchmark, not part of this initial comparison.
data_for_lme <- data_full %>%
  filter(Treatment %in% c("pbs", "propionate")) %>% 
  mutate(Treatment = factor(Treatment, levels = c('pbs','propionate')))

# Run the LME model for every pathway
lme_results <- data_for_lme %>%
  group_by(Pathway) %>%
  do({
    # The model tests if the change over time is different between the two treatment groups
    model <- lmer(log2_cpm ~ Timepoint * Treatment + (1 | MouseID), data = .)
    coeffs <- as.data.frame(summary(model)$coefficients)
    
    # Extract the result for the interaction term
    # This term directly tests our hypothesis for each pathway
    interaction_row <- coeffs[rownames(coeffs) == "Timepointpost:Treatmentpropionate", ] 
    
    data.frame(
      Estimate = interaction_row$Estimate,
      P_value = interaction_row$`Pr(>|t|)`
    )
  }) %>%
  ungroup()

#  Adjust P-values and Visualize ---

# Adjust for multiple testing using the Benjamini-Hochberg (FDR) method
lme_results_final <- lme_results %>%
  mutate(FDR = p.adjust(P_value, method = "BH")) %>%
  arrange(FDR)

# View top results
head(lme_results_final)


```
### volcano and abundance 

```{r}
# Define thresholds for significance and effect size
fdr_threshold <- 0.05
estimate_threshold <- 2.0

plot_data <- lme_results_final %>%
  mutate(
    # Create a column for coloring points based on significance and direction
    significance = case_when(
      FDR < fdr_threshold & Estimate > estimate_threshold  ~ "Increased",
      FDR < fdr_threshold & Estimate < -estimate_threshold ~ "Decreased",
      TRUE                                             ~ "Not Significant"
    ),
    # Create a column for labeling only the most interesting points
    label = ifelse(FDR < fdr_threshold & abs(Estimate) > estimate_threshold, Pathway, "")
  )

#  Create the Volcano Plot with ggplot2
volcano_plot <- ggplot(plot_data, aes(x = Estimate, y = -log10(FDR))) +
  # Add points, colored by our new 'significance' column
  geom_point(aes(color = significance), alpha = 0.7, size = 2.5) +
  
  # Add non-overlapping text labels for the top pathways using ggrepel
  geom_text_repel(aes(label = label), 
                  box.padding = 0.5, 
                  point.padding = 0.5,
                  max.overlaps = Inf, # Allow all labels to be plotted
                  segment.color = 'grey50',
                  size = 3.5) +
  
  # Define the custom colors for our groups
  scale_color_manual(values = c("Increased" = "#d62728", 
                                "Decreased" = "#1f77b4", 
                                "Not Significant" = "grey80")) +
  
  # Add vertical lines for the effect size (Estimate) threshold
  geom_vline(xintercept = c(-estimate_threshold, estimate_threshold), linetype = "dashed", color = "grey50") +
  
  # Add a horizontal line for the FDR significance threshold
  geom_hline(yintercept = -log10(fdr_threshold), linetype = "dashed", color = "grey50") +
  
  # Add labels and titles
  labs(
    title = "Propionate Treatment vs. PBS Control in transplant mice",
    subtitle = "Changes in Microbial Metabolic Pathways",
    x = "Log2 Fold Change (Estimate)",
    y = "-log10(FDR)",
    color = "Significance"
  ) +
  
  # Apply a clean, publication-ready theme
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
ggsave('../results/F4F.pdf', width = 8, height = 7)
```


```{r}
### the abundance interaction plot
top_pathways <- plot_data %>%
  filter(significance != 'Not Significant') %>% 
  arrange(desc(significance)) %>%       # Sort by FDR to get the most significant
  pull(Pathway)          # Extract the pathway names into a vector

#  Filter the main data frame for only these top pathways
plot_data <- data_full %>%
  filter(Pathway %in% top_pathways) %>%
  # Important for ordering the grid: convert Pathway to a factor based on significance
  mutate(Pathway = factor(Pathway, levels = top_pathways)) 

#  Create the faceted plot
# The core is the same, but we add `facet_wrap()` at the end.
faceted_interaction_plot <- ggplot(plot_data, 
                                   aes(x = Timepoint, y = log2_cpm, group = MouseID, color = Treatment)) +
  
  # Add lines for each mouse
  geom_line(alpha = 0.4, size = 0.8) +
  
  # Add points for each measurement
  geom_point(alpha = 0.7, size = 1.5) +
  
  # Add a thicker line showing the AVERAGE trend for each treatment group
  stat_summary(aes(group = Treatment), fun = mean, geom = "line", size = 1.5, linetype = 'dashed') +
  
  # Use consistent colors
  scale_color_manual(values = c("pbs" = "#1f77b4", "propionate" = "#d62728", "tcd" = "black")) +
  
  # *** THE KEY STEP: Create a grid of plots ***
  # `~ Pathway` tells ggplot to make a panel for each pathway.
  # `ncol = 4` arranges them into a grid with 4 columns.
  # `scales = "fixed"` ensures the Y-axis is the same for all panels, which is crucial for comparison.
  facet_wrap(~ Pathway, ncol = 3, scales = "fixed") +
  
  # Add informative labels and a title
  labs(
    title = "Significant Pathways: Pre vs. Post Treatment",
    subtitle = "Individual mouse trajectories shown, with group averages in bold",
    x = "Timepoint",
    y = "Log2(CPM + 1) Abundance",
    color = "Treatment Group"
  ) +
  
  # Apply a theme suitable for a faceted plot
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 8) # Style the facet titles
  )

faceted_interaction_plot
```

### the stratified output

```{r}
# We need to do some cleaning to separate the taxonomy string into levels
tax_raw <- read_tsv("../data/ASV_tax_assignments_transplant.txt", col_names = c('taxon','Taxon','X3','X4')) 
  
tax <- tax_raw %>%
  # Separate the long taxon string into taxonomic ranks
  separate(Taxon, into = c("k", "p", "c", "o", "f", "g", "s"), sep = ";") %>%
  # Clean up the names (e.g., remove "g__")
  mutate(g = str_remove(g, "g__")) %>%
  select(taxon, Genus = g) # Keep only the Genus level for this plot

# Load the large taxon contribution file
contrib <- read_tsv('../data/picrust2_out_transplant//pathways_out/path_abun_contrib.tsv.gz') %>% rename_with(~ "function_", .cols = 2)

#  Define Key Pathways and Prepare the Data ---

# Define the pathways we want to visualize
pathways_of_interest <- top_pathways

# Filter the contribution file for our pathways and merge with taxonomy and metadata
plot_data_prep <- contrib %>%
  filter(function_ %in% pathways_of_interest) %>%
  left_join(tax, by = "taxon") %>%
  # If a taxon has no genus assigned, label it "Unassigned"
  mutate(Genus = ifelse(Genus == 'NA' | Genus == "", "Unassigned", Genus)) %>%
  left_join(meta, by = c("sample" = "SampleID")) 


#  Summarize Data for Plotting ---
# The key value, taxon_function_abun, is calculated by multiplying the raw count of a microbe by its predicted gene copy number for that function. This gives you a single number representing that microbe's total functional contribution to a pathway in your sample.

# Find the top 10 most influential genera across all samples/pathways to keep the plot clean
top_genera <- plot_data_prep %>%
  group_by(Genus) %>%
  summarise(total_contrib = sum(taxon_function_abun)) %>%
  arrange(desc(total_contrib)) %>%
  slice_head(n =10) %>%
  pull(Genus)

# Summarize the data, grouping minor genera into an "Other" category
summary_data <- plot_data_prep %>%
  # Group everything not in our top list into "Other"
  mutate(Genus = ifelse(Genus %in% top_genera, Genus, "Other")) %>%
  # Group by our experimental variables and the new Genus column
  group_by(Treatment, Timepoint, Pathway = function_, Genus) %>%
  # Calculate the total contribution for each genus in each group
  summarise(avg_contribution = mean(taxon_function_abun), .groups = 'drop') %>%
  # Set the order of Timepoint for the plot
  mutate(Timepoint = factor(Timepoint, levels = c("pre", "post")))


# Create the Faceted Stacked Bar Plot ---

contribution_plot <- ggplot(summary_data, aes(x = Treatment, y = avg_contribution, fill = Genus)) +
  # Create the stacked bars
  geom_bar(stat = "identity", position = "stack") +
  
  # Split the plot into panels for each Timepoint (Pre vs. Post)
  facet_grid(Pathway ~ Timepoint, scales = "free_y") +
  
  # Use a color-blind friendly palette with enough colors for top genera + "Other"
  scale_fill_brewer(palette = "Paired") +
  
  # Add labels and titles
  labs(
    title = "Key Microbial Contributors to Metabolic Pathways in transplant mice",
    subtitle = "Comparing Pre- vs. Post-Treatment in PBS and Propionate Groups",
    x = "Treatment Group",
    y = "Average Pathway Abundance (Contribution)",
    fill = "Contributing Genus"
  ) +
   
  # Apply a clean theme
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "left",
    strip.text.x = element_text(face = "bold"),
    strip.text.y = element_text(face = "bold", size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Display the plot
print(contribution_plot)
ggsave('../results/01_transplant_3_pw_genus_F4G.pdf', width = 8, height = 7)
```
### PCOA visualizing the compositional change 

```{r}
# Load the principal coordinates file from QIIME 2.
pcoa_results_raw <- read_tsv("../data/weighted_unifrac_pc_transplant.txt", skip = 9, col_names = FALSE) %>% 
  mutate(X1 = tolower(str_replace_all(X1, '\\.','_')))

# The first column is the SampleID, the rest are the coordinates for each axis (PC1, PC2, etc.)
colnames(pcoa_results_raw) <- c("SampleID", paste0("PC", 1:(ncol(pcoa_results_raw)-1)))

# Extract the "Proportion Explained" ---

# We also want to know how much variance each axis explains. This is in the header.
proportion_explained_raw <- read_tsv("../data/weighted_unifrac_pc_transplant.txt", skip = 4, n_max = 1, col_names = FALSE)

# Extract the values for the first two axes and convert to percentage
pc1_varexp <- round(proportion_explained_raw$X1 * 100, 1)
pc2_varexp <- round(proportion_explained_raw$X2 * 100, 1)

#  Load Metadata and Merge ---

# Merge the coordinates with the metadata by SampleID
plot_data <- inner_join(pcoa_results_raw, meta, by = "SampleID") %>%
  # Make sure the groups are ordered logically for the plot legend
  mutate(
    Treatment = factor(Treatment, levels = c("pbs", "propionate", "tcd")),
    Timepoint = factor(Timepoint, levels = c("pre", "post"))
  ) 
```


```{r}
#  Create the PCoA Plot ---

pcoa_plot <- ggplot(plot_data, aes(x = PC1, y = PC2, color = Treatment, shape = Timepoint)) +
  # Add the points. We make them slightly transparent (alpha) and larger (size).
  geom_point(size = 4, alpha = 0.8) +
  
  # Use a clear, color-blind friendly color scheme
  scale_color_manual(values = c("pbs" = "#1f77b4", "propionate" = "#d62728", "tcd" = "black")) +
  
  # Use solid circles for 'pre' and solid triangles for 'post'
  scale_shape_manual(values = c("pre" = 16, "post" = 17)) +
  
  # Add lines connecting the pre- and post- samples for each mouse to show trajectory
  geom_line(aes(group = MouseID), alpha = 0.3) +

  # Add informative axis labels, including the proportion of variance explained
  labs(
    title = "PCoA of Microbial Communities (Weighted UniFrac)",
    subtitle = "Visualizing the trajectory of each treatment group",
    x = paste0("PC1 (", pc1_varexp, "%)"),
    y = paste0("PC2 (", pc2_varexp, "%)"),
    color = "Treatment Group",
    shape = "Timepoint"
  ) +
  
  # Apply a clean theme and add a coordinate system
  theme_minimal(base_size = 14) +
  coord_fixed() # Ensures that the axes are scaled 1:1

# Display the plot
print(pcoa_plot)
```
```{r}
# what about the other two distance matrix
# Load the principal coordinates file from QIIME 2.
pcoa_results_raw <- read_tsv("../data/bray_curtis_pc_transplant.txt", skip = 9, col_names = FALSE) %>% 
  mutate(X1 = tolower(str_replace_all(X1, '\\.','_')))

# The first column is the SampleID, the rest are the coordinates for each axis (PC1, PC2, etc.)
colnames(pcoa_results_raw) <- c("SampleID", paste0("PC", 1:(ncol(pcoa_results_raw)-1)))

# Extract the "Proportion Explained" ---

# We also want to know how much variance each axis explains. This is in the header.
proportion_explained_raw <- read_tsv("../data/bray_curtis_pc_transplant.txt", skip = 4, n_max = 1, col_names = FALSE)

# Extract the values for the first two axes and convert to percentage
pc1_varexp <- round(proportion_explained_raw$X1 * 100, 1)
pc2_varexp <- round(proportion_explained_raw$X2 * 100, 1)

#  Load Metadata and Merge ---

# Merge the coordinates with the metadata by SampleID
plot_data <- inner_join(pcoa_results_raw, meta, by = "SampleID") %>%
  # Make sure the groups are ordered logically for the plot legend
  mutate(
    Treatment = factor(Treatment, levels = c("pbs", "propionate", "tcd")),
    Timepoint = factor(Timepoint, levels = c("pre", "post"))
  ) 


pcoa_plot <- ggplot(plot_data, aes(x = PC1, y = PC2, color = Treatment, shape = Timepoint)) +
  # Add the points. We make them slightly transparent (alpha) and larger (size).
  geom_point(size = 4, alpha = 0.8) +
  
  # Use a clear, color-blind friendly color scheme
  scale_color_manual(values = c("pbs" = "#1f77b4", "propionate" = "#d62728", "tcd" = "black")) +
  
  # Use solid circles for 'pre' and solid triangles for 'post'
  scale_shape_manual(values = c("pre" = 16, "post" = 17)) +
  
  # Add lines connecting the pre- and post- samples for each mouse to show trajectory
  geom_line(aes(group = MouseID), alpha = 0.3) +

  # Add informative axis labels, including the proportion of variance explained
  labs(
    title = "PCoA of Microbial Communities (Bray Curtis)",
    subtitle = "Visualizing the trajectory of each treatment group",
    x = paste0("PC1 (", pc1_varexp, "%)"),
    y = paste0("PC2 (", pc2_varexp, "%)"),
    color = "Treatment Group",
    shape = "Timepoint"
  ) +
  
  # Apply a clean theme and add a coordinate system
  theme_minimal(base_size = 14) +
  coord_fixed() # Ensures that the axes are scaled 1:1

# Display the plot
print(pcoa_plot)
```

```{r}
#  Statistical Testing with PERMANOVA ---
# This test will tell us if the group centroids are statistically different,
# even if the points overlap visually.

# Load the distance matrix
dist_matrix_raw <- read_tsv("../data/bray_curtis_dm_transplant.txt") 

cleaned_colnames <- tolower(str_replace_all(colnames(dist_matrix_raw)[-1], "\\.", "_"))

dist_matrix_clean <- dist_matrix_raw %>% 
  # Clean the first column
  mutate(across(1, ~ tolower(str_replace_all(.x, "\\.", "_")))) %>%
  # Set that cleaned first column as the row names
  column_to_rownames(var = colnames(dist_matrix_raw)[1])

# Assign the cleaned column names
colnames(dist_matrix_clean) <- cleaned_colnames

# Convert the cleaned data frame into a 'dist' object for the vegan package
dist_obj <- as.dist(dist_matrix_clean)

# It's critical that the metadata rows are in the same order as the distance matrix.
# This code reorders the metadata to match the distance matrix labels perfectly.
meta_for_permanova <- meta %>%
  dplyr::slice(match(labels(dist_obj), SampleID)) 
```


```{r}
# Run the PERMANOVA test
# We use the formula `dist_matrix ~ Timepoint * Treatment` to test the interaction.
# This asks: "Is the trajectory of community change different between the treatment groups?"

# adonis2() gets its p-values by randomly permuting the group labels, so the result
# differs slightly on every run unless the generator is seeded. Seed here as well as
# in the setup chunk, so that re-running this one chunk reproduces the published
# p-values exactly rather than only a full top-to-bottom knit.
set.seed(20250602)
permanova_result <- adonis2(dist_obj ~ Timepoint * Treatment, data = meta_for_permanova, by = "terms")

# Display the results
print(permanova_result)
```

### Identify the "Propionate-Signature" Bacteria 

```{r}
# --- Load and Format Data ---

# Load the ASV count data from the exported .biom file
# The file path will be inside the folder you exported to.
count_table <- read_csv("../data/transplant_ASV_Abundance_Table.csv" ) %>%  column_to_rownames('seqs')
# Convert the biom object to a standard numeric matrix
count_matrix <- as.matrix(count_table) %>% t()

colnames(count_matrix) <- tolower(str_replace_all(colnames(count_matrix), "\\.", "_"))

# Load the taxonomy data
tax_raw <- read_tsv("../data/ASV_tax_assignments_transplant.txt", col_names = F)
tax_clean <- tax_raw %>%
  dplyr::rename(ASV = X1) %>%
  # Separate the long taxon string into ranks for easier labeling later
  separate(X2, into = c("k", "p", "c", "o", "f", "g", "s"), sep = ";") %>%
  # Clean up the names (e.g., remove "g__")
  mutate(g = str_remove(g, "g__"),
         s = str_remove(s, "s__"),
         f = str_remove(f, "f__")) %>%
  select(ASV, Family = f, Genus = g, Species = s)

# --- Prepare Data for DESeq2 ---

# The goal is to find the signature bacteria AFTER treatment.
# So, we will filter our data to only include the 'post' samples.
meta_post <- meta %>%
  mutate(Timepoint = as.character(Timepoint)) %>% 
  filter(Timepoint == "post")

# The columns in the count matrix must be in the same order as the rows in the metadata.
# This reorders the count matrix columns to match the metadata.
count_matrix_post <- count_matrix[, meta_post$SampleID]

# --- Run DESeq2 Analysis ---

# Create the DESeqDataSet object. This object bundles the counts,
# metadata, and the experimental design for the analysis.
# The design `~ Treatment` tells DESeq2 to find differences based on the Treatment column.
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix_post,
  colData = meta_post,
  design = ~ Treatment
)

# Run the main DESeq2 function. This performs normalization, dispersion estimation,
# and fits the model to find differentially abundant features.
dds_result <- DESeq(dds)

# Get the results for our comparison of interest: propionate vs. the PBS control.
# A positive log2FoldChange means the ASV is more abundant in the propionate group.
res <- results(dds_result, contrast = c("Treatment", "propionate", "pbs"))

# Convert the results to a tidy data frame for plotting
res_df <- as.data.frame(res) %>%
  rownames_to_column(var = "ASV") %>%
  # Add the taxonomy information to our results
  left_join(tax_clean, by = "ASV") %>%
  # Filter for statistically significant results (FDR < 0.05)
  filter(padj < 0.05) %>%
  # Sort by the fold change to see the biggest differences first
  arrange(desc(log2FoldChange))

# --- Visualize the Results ---

# Create a clean label for the plot, using Genus and ASV ID.
res_df <- res_df %>%
  mutate(PlotLabel = str_glue('f_{Family}|g_{Genus}|s_{Species} ({ASV})'))

# Create a "dot plot" of the top 20 signature ASVs
signature_plot <- ggplot(head(res_df, 20), aes(x = log2FoldChange, y = reorder(PlotLabel, log2FoldChange))) +
  geom_point(aes(color = log2FoldChange, size = -log10(padj)), alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_gradient2(low = "#1f77b4", mid = "grey", high = "#d62728", midpoint = 0) +
  labs(
    title = "Top Propionate-Signature Bacteria (vs. PBS)",
    subtitle = "Analysis of post-treatment samples",
    x = "Log2 Fold Change (More abundant in Propionate ->)",
    y = "ASV (annotated with Genus)",
    color = "Log2 Fold Change",
    size = "Significance (-log10 FDR)"
  ) +
  theme_minimal()

# Display the plot
print(signature_plot)
ggsave('../results/F4E.pdf', width = 8, height = 7)
```

### correlate with the gvhd score

```{r}
# --- Step 1: Read the clinical GVHD scores from their source spreadsheet ---

# These scores used to be typed into the script by hand. They are now read straight
# from data/Exp85 Propionate GVHD scores.xlsx so the numbers cannot drift from the
# source, and so it is obvious where they came from.
#
# The sheet is laid out as a scoring grid rather than a tidy table: row 1 holds the
# group name above the first mouse of each group, row 2 holds the mouse numbers, and
# each row from 3 down is one scoring day. Blank spacer columns separate the groups.
gvhd_raw <- read_excel("../data/Exp85 Propionate GVHD scores.xlsx",
                       sheet = 1, col_names = FALSE, .name_repair = "minimal")

# Row 2 carries the mouse numbers. Column 1 is the day label, and any column with no
# mouse number is a spacer, so those are the columns to drop.
mouse_numbers <- as.character(unlist(gvhd_raw[2, ]))
score_cols    <- which(!is.na(mouse_numbers))
score_cols    <- score_cols[score_cols != 1]

# Row 1 names a group only above its first mouse, so carry each name rightward to
# cover that group's remaining columns.
group_labels <- as.character(unlist(gvhd_raw[1, ])) %>%
  # tidyr::fill() works on columns, so pivot through a one-column tibble
  {tibble(g = .)} %>%
  fill(g, .direction = "down") %>%
  pull(g)

# Normalize the sheet's headings ("TCD group ", "PO PBS ", "PO Propionate") to the
# lowercase treatment codes used everywhere else in this notebook.
treatment_of <- function(x) {
  x <- str_squish(str_to_lower(x))
  case_when(str_detect(x, "propionate") ~ "propionate",
            str_detect(x, "pbs")        ~ "pbs",
            str_detect(x, "tcd")        ~ "tcd",
            TRUE                        ~ NA_character_)
}

# Assemble one row per mouse per scoring day, then keep day 7 — the timepoint the
# stool samples were collected at, and the one reported in the figure.
gvhd_scores_day7 <- tibble(
  col         = score_cols,
  Treatment   = treatment_of(group_labels[score_cols]),
  MouseNumber = as.integer(mouse_numbers[score_cols])
) %>%
  mutate(GVHD_Score = map_dbl(col, ~ as.numeric(gvhd_raw[[.x]][3]))) %>%
  # Row 3 of the sheet is day 7; confirm that before trusting it
  {stopifnot(as.numeric(gvhd_raw[[1]][3]) == 7); .} %>%
  select(Treatment, MouseNumber, GVHD_Score) %>%
  # MouseID must match the format built from the sample names in the metadata
  mutate(MouseID = paste(Treatment, MouseNumber, sep = "_"))

# Guard against a silent change in the spreadsheet layout: these are the day 7
# scores as published, so any mismatch means the parsing above needs revisiting.
stopifnot(
  nrow(gvhd_scores_day7) == 20,
  identical(gvhd_scores_day7$GVHD_Score,
            c(0, 0, 0, 0,                            # TCD
              5, 3.5, 4, 4, 3.5, 3.5, 4, 4,          # PBS
              2.5, 2.5, 2.5, 2, 2, 2.5, 2, 2))       # propionate
)


# --- Step 2: Prepare the Microbiome and Clinical Data for Merging ---

# `data_full` is reused between the steady-state and transplant sections of this
# notebook. Assert we are holding the transplant version, so that running this chunk
# out of order fails loudly instead of quietly analyzing the wrong experiment.
stopifnot(all(c("pre", "post") %in% as.character(data_full$Timepoint)),
          all(c("pbs", "propionate", "tcd") %in% data_full$Treatment))

# Choose the specific microbial feature we want to test.
# Let's test the L-citrulline biosynthesis pathway (CITRULBIO-PWY)
pathway_to_test <- "CITRULBIO-PWY"

# Filter the main microbiome data for our pathway of interest and the 'post' timepoint.
pathway_data_post <- data_full %>%
  mutate(Timepoint = as.character(Timepoint)) %>%
  filter(Timepoint == "post", Pathway == pathway_to_test)

# Now, join this microbiome data with the corresponding GVHD scores.
# We use an inner_join on MouseID to link each mouse's pathway abundance to its GVHD score.
correlation_data <- inner_join(pathway_data_post, gvhd_scores_day7, by = "MouseID")

# Every mouse must match; a silent drop here would change the statistics below.
stopifnot(nrow(correlation_data) == nrow(gvhd_scores_day7))


# --- Step 3: Quantify the group separation ---

# NOTE ON METHOD (2026-08): this section previously reported a Spearman rank
# correlation between pathway abundance and GVHD score, which dated from an earlier
# version of the figure that drew a regression line within each group. That test does
# not support the claim being made and is not significant in any formulation:
#
#   all three groups pooled : rho = +0.06, p = 0.79
#   PBS + propionate only   : rho = -0.49, p = 0.053
#   within PBS only         : rho = -0.21, p = 0.62
#   within propionate only  : rho = +0.44, p = 0.28
#
# Pooling turns the correlation positive because TCD mice have both low pathway
# abundance and a GVHD score of 0, which cancels the trend in the other two groups.
#
# The claim the figure actually makes is that propionate-treated mice sit in a
# distinct region of high pathway abundance and low disease score, so the reported
# statistics are now group differences on each axis. Those are unambiguous.
group_tests <- correlation_data %>%
  filter(Treatment.x != "tcd") %>%
  # Wilcoxon rank-sum: non-parametric, appropriate for n = 8 per group and for the
  # ordinal, heavily tied GVHD score
  summarise(
    p_pathway = wilcox.test(log2_cpm   ~ Treatment.x, exact = FALSE)$p.value,
    p_gvhd    = wilcox.test(GVHD_Score ~ Treatment.x, exact = FALSE)$p.value
  )

print(group_tests)

# Group means, which are what the figure legend quotes
correlation_data %>%
  group_by(Treatment.x) %>%
  summarise(n = n(),
            mean_log2_cpm = mean(log2_cpm), sd_log2_cpm = sd(log2_cpm),
            mean_gvhd     = mean(GVHD_Score), sd_gvhd    = sd(GVHD_Score)) %>%
  print()

# The superseded correlation, kept so the numbers above can be checked against the
# earlier version of the analysis. It is not reported in the manuscript.
correlation_test <- cor.test(~ log2_cpm + GVHD_Score,
                             data = correlation_data,
                             method = "spearman", exact = FALSE)
print(correlation_test)
```


```{r}
# --- Step 4: Visualize the Group Separation  ---

# This plot emphasizes the distinct clustering of the treatment groups.
separation_plot <- ggplot(correlation_data, aes(x = log2_cpm, y = GVHD_Score)) +
  # --- EDIT ---
  # Add ellipses for only the PBS and Propionate groups to avoid the error.
  # We do this by passing a filtered dataset directly to this layer.
  stat_ellipse(data = . %>% filter(Treatment.x != "tcd"), 
               aes(color = Treatment.x), 
               type = "t", 
               level = 0.95, 
               linewidth = 1.5, 
               alpha = 0.7) +
  
  # Add the individual data points.
  geom_point(aes(color = Treatment.x), size = 4, alpha = 0.8) +
  
  # Use our standard color scheme.
  scale_color_manual(values = c("pbs" = "#1f77b4", "propionate" = "#d62728", "tcd" = "black")) +
  
  # Add informative labels that focus on the separation.
  labs(
    title = paste("Separation of Treatment Groups by", pathway_to_test, "and GVHD Score"),
    subtitle = "Propionate treatment shifts mice to a 'high pathway, low disease' state",
    x = "Pathway Abundance (Log2 CPM)",
    y = "GVHD Score at Day +7",
    color = "Treatment Group"
  ) +
  
  theme_minimal(base_size = 14)

# Display the plot
print(separation_plot)

# Write to results/ alongside the other figure panels. The original version wrote
# this PDF into data/, which mixed a generated figure in with the raw inputs.
ggsave('../results/F4H.pdf', plot = separation_plot, width = 8)
```

```{r}
# Record the exact package versions this run used, so a future reader can tell
# whether a difference in output comes from the code or from the environment.
sessionInfo()
```

